### 2022-02-11 chapters/intro/intro.tex
finished hydro part in intro

\section{Modeling the hydrodynamics of the FRT}
The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean. The large-scale circulation in this region is dominated by the interplay of the Florida Current (FC) and the Loop Current (LC). The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC). The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream. The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}. The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT. These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability. The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}. These eddies significantly impact the ecology of the southwest Florida Shelf as they have been identified as a mechanism causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}. Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}. These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}. On the northernmost section of the FRT however, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}. The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}

@@

The aforementioned large-scale ocean circulation significantly impact transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. Unstructured-mesh models offer a well-suited solution to tackle the multi-scale nature of flow features in the FRT.

@@

\section*{Supporting publications}

>>>>

\section{Modeling hydrodynamics and transport processes in the FRT}
The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean. The large-scale circulation in this region is dominated by the interplay of the Florida Current (FC) and the Loop Current (LC). The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC). The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream. The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}. The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT. These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability. The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}. These eddies significantly impact the ecology of the southwest Florida Shelf as they have been identified as a mechanism causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}. Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}. These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}. On the northernmost section of the FRT however, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}. The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}

@@

The aforementioned large-scale ocean circulation significantly impact transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. Unstructured-mesh models offer a well-suited solution to cature the multi-scale nature of flow features in the FRT in a computationally tractable way. They allow to locally increase the model resolution close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}. Nested grid models are another possible approach \citep{warner2010development} but they lack the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}.

In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}). This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef \citep{lambrechts2008multi}. Moreover, by coupling SLIM with the HYbrid Coordinate Ocean Model (HYCOM-GOMl0.04, \citealp{chassignet2007hycom}), SLIM was able to indirectly represent the baroclinic dynamics of the FC. Using this coupling, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT, \ie~the exchanges of larvae between reefs. In this study, coral larvae were modeled using a Lagrangian particle tracking model forced by SLIM currents. To answer our different research questions, we adapted this model to evaluate the transport of the causative agent of the SCTLD and sediments particles.

However, at the beginning this thesis, SLIM was not able to represent wave-current interactions under storm conditions. Heavy winds generate large wind-waves that disturb ocean conditions and interact non-linearly with currents \citep{liu2020impacts,wu2011fvcom}. The wave-generated currents cause water level variations near shorelines and wave breaking points which, in turn, affect the motion and evolution of the waves \citep{longuet1970longshore, sikiric2013coupling}. Capturing wave-current interactions hence requires the calculation of the full directional wave spectrum in order to correctly reproduce the dynamics of wind-driven surface waves. This is usually achieved by spectral wave models, which describe the evolution of the wave energy spectrum. An objective of this thesis was therefore the coupling of SLIM with the spectral wave model Simulating WAve Nearshore (SWAN, \citealp{booij1999third}) SWAN was specifically developed for coastal applications and relies on numerical schemes adapted to small-scale, shallow water regions.

@@

\newpage
\section*{Supporting publications}

### 2022-02-11 include/title.tex
slim+swan paper added
chapter on drto added to thesis
draft onset added to manuscript
finished hydro part in intro

\vspace*{1.2cm}
\hspace{-0.8cm}
\begin{minipage}{1.02\linewidth}
\centering
\LARGE\bfseries How modeling tools can inform coral reef management in the Florida Reef Tract
\end{minipage}

\vspace*{0.5cm}
\hspace{-0.5cm}\begin{minipage}{\textwidth}
\centering
\rule{50pt}{.5pt}\\[1.2ex]
Doctoral dissertation presented by \\ [2.ex]
{\Large Thomas Dobbelaere\\ [2.ex]
}
in partial fulfillment of the requirements for the degree of\\ [2.ex]
\Large Docteur en Sciences de l'ingénieur

@@

\vspace*{0.5cm}
\footnotesize{
\hspace{-1.2cm}
\begin{tabular}{ll}
Prof. Emmanuel Hanert (Supervisor) & Université catholique de Louvain \\
Prof. Eric Deleersnijder           & Université catholique de Louvain \\
Dr. Jonathan Lambrechts            & Université catholique de Louvain\\
???                                & ??? \\
???                                & ??? \\ 
\end{tabular}
}
\end{center}
\end{minipage}

>>>>

\vspace*{1.2cm}
\hspace{-0.8cm}
\begin{minipage}{1.02\linewidth}
\centering
\LARGE\bfseries Better understanding transport processes to support Florida's Coral Reef % How modeling tools can inform coral reef management in the Florida Reef Tract
\end{minipage}

\vspace*{0.5cm}
\hspace{-0.5cm}\begin{minipage}{\textwidth}
\centering
\rule{50pt}{.5pt}\\[1.2ex]
Doctoral dissertation presented by \\ [2.ex]
{\Large Thomas Dobbelaere\\ [2.ex]
}
in partial fulfillment of the requirements for the degree of\\ [2.ex]
\Large Doctor in Engineering Science

@@

\vspace*{0.5cm}
\footnotesize{
\hspace{-1cm}
\begin{tabular}{ll}
Prof. Emmanuel Hanert (Supervisor) & Université catholique de Louvain \\
Prof. Eric Deleersnijder           & Université catholique de Louvain \\
Dr. Jonathan Lambrechts            & Université catholique de Louvain\\
Prof. Claire B. Paris-Limouzy      & University of Miami \\
Dr. Christopher Thomas             & Université catholique de Louvain \footnote{Present address: Risk Management Solutions, London, UK}
\end{tabular}
}
\end{center}
\end{minipage}
